Particle Smoothing Algorithms for Variable Rate Models

### 2012-04-20 full_finaldraft_VRsmoothing_withSJGcomments.tex
Simon's mods

\title{Smoothing Algorithms for Variable Rate Models}

@@

% The paper headers
\markboth{IEEE Transaction in Signal Processing,~Vol.~1, No.~1, January~1901}%
{Bunch \& Godsill: Smoothing Algorithms for Variable Rate Models}

@@

The VRPF allows the changepoint sequence -- and hence the current state -- to be estimated sequentially as observations are received. However, estimates can often be improved later once further observations have been made. In this paper, we address the problem of smoothing in variable models, i.e. the estimation of the changepoint sequence and latent state given all the observations. This is achieved with an efficient backward sweep through the observations, in a similar manner to the method for standard HMMs described in \cite{Godsill2004}. Two new schemes are introduced: one for conditionally linear-Gaussian models which exploits the method of Rao-Blackwellisation; and a second for use with conditionally-deterministic models which uses an augmented target distribution in the style of an SMC sampler \cite{DelMoral2006}.

@@

We consider a general model from time $0$ to $T$, between which observations, $\{y_1 \dots y_N\}$, are made at times $\{t_1 \dots t_N = T\}$. During this period, an unknown number of changepoints, $K$, occur at times $\{\tau_0 = 0, \tau_1 \dots \tau_K \}$, each with associated changepoint parameters, $\{ u_0, u_1 \dots u_K \}$. The pairs $\{\tau_k, u_k\}$ are the elements of an marked point process (MPP). The latent state is a continuous-time process denoted $x(t)$. Discrete sets containing multiple values over time will be written as, e.g. $y_{1:n} = \{y_1 \dots y_n\}$.

@@

The changepoint sequence is assumed to be a Markov process.
%
\begin{IEEEeqnarray}{rCl}
 \{\tau_k, u_k\} & \sim & p(\tau_k, u_k|\tau_{k-1}, u_{k-1}) \label{eq:cp_model}
\end{IEEEeqnarray}

@@

@@

The first class of variable rate models to be considered is those whose state dynamics are linear-Gaussian conditional on the changepoint sequence. Such a model may be discretised onto the set of observation times in exactly the same manner as a standard the standard HMM.
%
\begin{IEEEeqnarray}{rCl}
 x_n & = & A_n(\theta_{n})x_{n-1} + w_n \\
 y_n & = & C_n(\theta_{n})x_n + v_n       ,
\end{IEEEeqnarray}

where,
%
\begin{IEEEeqnarray}{rCl}
 w_n & \sim & \mathcal{N}(w_n|0, Q_n(\theta_{n})) \\
 v_n & \sim & \mathcal{N}(v_n|0, R_n(\theta_{n}))       .
\end{IEEEeqnarray}

In addition, the prior state distribution should be Gaussian, with known mean and variance.
%
\begin{IEEEeqnarray}{rCl}
 x_0 & \sim & \mathcal{N}(x_0|m_0, P_0)       .
\end{IEEEeqnarray}

If the changepoint sequence is known, or has been estimated, then the state values, $x_n$ may be inferred using optimal Kalman filtering and smoothing recursions. As well as the basic Kalman filter \cite{Kalman1960}, the Rauch-Tung-Striebel (RTS) smoother \cite{Rauch1965} and two-filter smoother \cite{Fraser1969} will prove useful for this step.

Conditionally linear-Gaussian variable rate models were introduced in \cite{Godsill2007a} for a financial inference algorithm. Changepoints correspond to jumps in the value or trend of a security, at which points the process covariance, $Q_n(\theta_n)$, is inflated.

@@

In general, the state dynamics will be governed by a differential equation which may depend on the entire changepoint sequence. Here we assume that only the most recent changepoint is significant.
%
\begin{IEEEeqnarray}{rCl}
 dx(t) & = & \mathrm{f}(x(t), \tau_{K(t)}, u_{K(t)})     .
\end{IEEEeqnarray}

By introducing a new sequence, $\{ x_0, x_1 \dots x_K \}$, which denotes the value of the state at each changepoint (i.e. $x(\tau_k)$), and assuming that an analytic solution exists, a state transition function may be found,
%
\begin{IEEEeqnarray}{rCll}
 x(t) & = & f(x_{K_n}, u_{K_n}, \tau_{K_n}, t) &, \tau_{K_n} < t \leq \tau_{K_{n}+1}    \label{eq:disc_time_state_diff_eq}     .
\end{IEEEeqnarray}

By choosing $t = \tau_{K_{n}+1}$, this equation specifies the state at the next changepoint time. Similarly, by choosing $t=t_n$, the state at the observation times may be calculated --- these points will be denoted $\hat{x}_n$. To complete the model, a probabilistic measurement model must be devised for the observation process, $p(y_n|\hat{x}_n)$.

@@

The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1})$, has a similar form to the changepoint prior of (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}. An additional survivor function is included to account for the condition that changepoints cannot occur before $t_{n-1}$.
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1}) = S(\tau_{K_n}, t_n) / S(\tau_{K_{n-1}+1}, t_{n-1}) } \nonumber \\
 \qquad  \qquad & & \times \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1})  \label{eq:cp_sequence_trandens}
\end{IEEEeqnarray}
%\begin{IEEEeqnarray}{rCl}
%\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1})} \nonumber \\
%    & = & p(\tau_{K_{n-1}+1:K_n}, u_{K_{n-1}+1:K_n}, \tau_{K_n+1}>t_n|\tau_{K_{n-1}+1}>t_{n-1}, \tau_{K_{n-1}}, u_{K_{n-1}}) \nonumber \\
%    & = & \begin{cases} S(\tau_{K_n}, t_n) \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1}, \tau_j>t_{n-1}) & K_n > K_{n-1} \\
%                        S(\tau_{K_n}, t_n) / S(\tau_{K_n}, t_{n-1}) & K_n = K_{n-1} \end{cases} \IEEEeqnarraynumspace \label{eq:cp_sequence_trandens}
%\end{IEEEeqnarray}

@@

Practically, because changepoints will be relatively rare events, it is not likely that more than one new changepoint will occur between $t_{n-1}$ and $t_n$.

The target distribution of (\ref{eq:vrpf_target}) cannot be calculated analytically, but may be approximated numerically. A particle filter is an algorithm for approximating a probability distribution using a set of weighted samples (or ``particles'') drawn from that distribution. In this case, each particle will be a set of changepoint times and parameters.
%
\begin{equation}
 \hat{p}(\theta_{n}|y_{1:n}) = \sum_j w_n^{(j)} \delta_{\theta_{n}^{(j)}}(\theta_{n})
\end{equation}

@@

The particle filter works recursively. At the $n$th step, a particle, $\theta_{n-1}^{(i)}$, is first resampled from those approximating the filtering distribution at the $(n-1)$th step, using an appropriately chosen set of proposal weights.
%
\begin{equation}
 q(\theta_{n-1}) = \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})
\end{equation}

@@

Next, an extension to the changepoint sequence, $\theta_{n \setminus n-1}^{(i)}$, is proposed from an importance distribution, $q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n)$, and concatenated with $\theta_{n-1}$ to create an estimate of $\theta_n$. Finally, the particle is weighted according to the ratio of the target and proposal densities.
%
\begin{IEEEeqnarray}{rCl}
w_n^{(i)} & = & \frac{ p(\theta_{n}^{(i)}|y_{1:n}) }{ q(\theta_{n}) } \nonumber \\
    & \propto & \frac{ p(y_n|\theta_n, y_{1:n-1}) p(\theta_{n \setminus n-1}|\theta_{n-1}) p(\theta_{n-1}|y_{1:n-1}) }{ q(\theta_{n-1}) q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n) } \nonumber \\
    & =       & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times \frac{ p(y_n|\theta_n, y_{1:n-1}) p(\theta_{n \setminus n-1}|\theta_{n-1}) }{ q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n) } \label{eq:vrpf_weights}
\end{IEEEeqnarray}

@@

For the most basic ``bootstrap'' \cite{Gordon1993} form of the VRPF, $\theta_{n \setminus n-1}$ may be proposed from the prior transition density (\ref{eq:cp_sequence_trandens}). This can be achieved by sampling new changepoints sequentially from the transition model (\ref{eq:cp_model}) until one falls after the current time, $t_n$. This final future changepoint is discarded. (This process can be thought of as sampling the entire future changepoint sequence from $t_{n-1}$ onwards, and then marginalising those which fall after $t_n$.) The bootstrap proposal leads to the usual simplification of the weight formula. % (apart from the first which is sampled from (\ref{eq:cp_cond_model}))
%
\begin{IEEEeqnarray}{rCl}
w_n^{(i)} & = & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times p(y_n|\hat{x}_n) \label{eq:bootstrap_vrpf_weights}
\end{IEEEeqnarray}

@@

The bootstrap versions of the VRPF may perform poorly if changepoints are not obvious until significantly after they occur. For example, in a tracking example, if a jump occurs in the acceleration, but only the position is observed, then this change may not be obvious until a number of observations have arrived. In this case, the estimation may be improved by the introduction of resample-move (RM) steps \cite{Gilks2001}. In an RM scheme, optional Metropolis-Hastings (MH) moves are conducted to alter the particle states after the importance sampling has taken place. For variable rate models, any one of the previous changepoints, $\tau_k$, or associated parameters, $u_k$, could be adjusted. Because more observations are available than when the changepoint was first proposed, it may be possible to construct more informed proposals and so move the changepoints towards regions with higher posterior probability. It is even possible to retrospectively add or remove changepoints, using reversible jump MH moves \cite{Green1995}. Variable rate particle filters using RM with piecewise deterministic models are described in \cite{Whiteley2011,Gilholm2008}.

@@

A filter conducts inference sequentially as new observations are introduced. The purpose of a smoother is to produce a revised estimate once all the observations have been made, using future values to improve upon the filter performance. Estimating changepoints online is a challenging task because the presence of a change may not be obvious until after it has happened. Thus, a smoothing algorithm is expected to achieve improved performance at changepoint estimation.

The target distribution for a variable rate smoothing algorithm is the posterior distribution over the entire changepoint sequence, $p(\theta|y_{1:N})$. This distribution is approximated by the final step of the VRPF. However, in the same manner as the fixed rate filter-smoother of \cite{Kitagawa1996}, this approximation is likely to lack path-space diversity --- because of the necessary resampling step in the filtering algorithm, the particles all share the same set of changepoints before a particular time, with variation only appearing for changepoints closer to $T$. For a better characterisation of the smoothing distribution, it is necessary to rejuvenate the set of particles. This is achieved with a backward pass through the observations in a similar manner to the forward-backward method described in \cite{Godsill2004}.

@@

It remains to devise a scheme for sampling the conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$. The simplifications used by the fixed rate particle smoother which exploit the Markovian nature of the state sequence \cite{Godsill2004} are of no help, because past changepoints are not independent of future observations (i.e. $p(\theta_{n}|\theta_{n}^{+}, y_{1:N}) \ne p(\theta_{n}|\theta_{n}^{+}, y_{1:n}))$. However, when the state dynamics are linear-Gaussian conditional on the changepoint sequence, it is possible split the dependence on the observations into past and future by introducing the current state as an additional variable. This trick was used in \cite{Sarkka2012} in the derivation of the Rao-Blackwellised particle filter for fixed rate conditionally linear-Gaussian models.
%
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{p(\theta_{n}|\theta_{n}^+, y_{1:N})} \nonumber \\
\qquad & \propto & p(\theta_{n}, \theta_{n}^+| y_{1:N}) \nonumber  \\
       & =       & \int p(x_n, \theta_{n}, \theta_{n}^+| y_{1:N}) dx_n \nonumber  \\
       & \propto & \int p(y_{n+1:N}|x_n, \theta_{n}, \theta_{n}^+, y_{1:n}) p(x_n, \theta_{n}, \theta_{n}^+| y_{1:n}) dx_n \nonumber \\
       & = & \int p(y_{n+1:N}|x_n, \theta_{n}^+) p(x_n|\theta_{n}, y_{1:n}) dx_n \nonumber \\
       &   & \times p(\theta_{n}^+|\theta_{n}) p(\theta_{n}|y_{1:n})
\end{IEEEeqnarray}

Finally, the particle approximation is substituted for the filtering distribution.
%
\begin{IEEEeqnarray}{rCl}
\hat{p}(\theta_{n}|\theta_{n}^+, y_{1:N}) & = & \sum_i \tilde{w}_{n}^{(i)} \delta_{\theta_{n}^{(i)}}(\theta_{n})  \label{eq:rb-vrps_back_cond}
\end{IEEEeqnarray}

where the backwards conditional weights are given by
%
\begin{IEEEeqnarray}{rCl}
 \tilde{w}_n & \propto & w_n \int p(y_{n+1:N}|x_n, \theta_{n}^+) \nonumber \\
             &         & \times p(x_n|\theta_{n}^{(i)}, y_{1:n}) dx_n p(\theta_{n}^+|\theta_{n}^{(i)}) \label{eq:rb-vrps_back_cond_weight}
\end{IEEEeqnarray}

@@

The integral in~(\ref{eq:rb-vrps_back_cond_weight}) contains two terms involving the current state, $x_n$. The first is the familiar state posterior generated by the Kalman filter using (\ref{eq:kf_predict_start}--\ref{eq:kf_update_stop}).
%
\begin{IEEEeqnarray}{rCl}
p(x_n|\theta_{n}^{(i)}, y_{1:n}) & = & \mathcal{N}(x_n|m_n^{(i)}, P_n^{(i)})
\end{IEEEeqnarray}

@@

The second state term in (\ref{eq:rb-vrps_back_cond_weight}) is the likelihood, $p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)$. This may be considered to be an improper density over $x_n$, and may be calculated analytically using a backwards Kalman filter, in a similar manner to that used in the two-filter smoother \cite{Fraser1969,Kitagawa1994,Sarkka2012,Gelb1974}. Such a backwards Kalman filter uses the following recursions, which are derived in appendix~\ref{app:backwards_KF}.
%
\begin{IEEEeqnarray}{rCl}
 p(y_{n+1:N}|x_n, \theta_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \\
 p(y_{n:N}|x_n, \theta_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n)
\end{IEEEeqnarray}
%
\begin{IEEEeqnarray}{rCl}
 \tilde{m}_n^- & = & A_{n+1}^{-1} \tilde{m}_{n+1} \label{eq:backward_kf_predict_start} \\
 \tilde{P}_n^- & = & A_{n+1}^{-1} (\tilde{P}_{n+1} + Q_{n+1}) A_{n+1}^{-T} \label{eq:backward_kf_predict_stop} \\
 \tilde{\mu}_n & = & C_n \tilde{m}_n^- \label{eq:backward_kf_update_start} \\
 \tilde{S}_n   & = & C_n \tilde{P}_n^- C_n^T + R_n \\
 \tilde{K}_n   & = & \tilde{P}_n^- C_n^T \tilde{S}_n^{-1} \\
 \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{K}_n (y_n - \tilde{\mu}_n) \\
 \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{K}_n \tilde{S}_n \tilde{K}_n^T \label{eq:backward_kf_update_stop}
\end{IEEEeqnarray}

@@

Substituting into (\ref{eq:rb-vrps_back_cond_weight}), the backwards conditional weights are given by,
%
\begin{equation}
 \tilde{w}_n \propto w_n p(\theta_{n}^+|\theta_{n}^{(i)}) \mathcal{N}(\tilde{m}_n^-|m_n, \tilde{P}_n^- + P_n)     .
\label{eq:rb-vrps_back_cond_weight2}
\end{equation}

Using these weights, a sample of $\theta_{n}$ may be drawn from the particle distribution of (\ref{eq:rb-vrps_back_cond}), completing the smoothing algorithm. Once sampling has progressed backwards from $n=N \dots 1$, a particle from the smoothing distribution will have been generated. This procedure may then be repeated until sufficient particles have been obtained. The algorithm is summarised in Fig.~\ref{alg:RBVRPS}.

@@

In a conditionally deterministic system, the entire continuous time state trajectory, $x(t)$, is a function of the changepoint sequence $\theta$ and the initial state, $x_0$. When the changepoint sequence is split into past, $\theta_n$, and future $\theta_n^+$, and the past is altered, this will alter the state trajectory in the future. This is undesirable, as each sampling operation will require the recalculation of state values and observations likelihoods for all future observation times.

>>>>

\title{Particle Smoothing Algorithms for Variable Rate Models}

@@

% The paper headers
\markboth{IEEE Transaction in Signal Processing,~Vol.~1, No.~1, January~1901}%
{Bunch \& Godsill: Particle Smoothing Algorithms for Variable Rate Models}

@@

The VRPF allows the changepoint sequence -- and hence the current state -- to be estimated sequentially as observations are received. However, estimates can often be improved later once further observations have been made. In this paper, we address the problem of smoothing in variable rate models, i.e. the estimation of the changepoint sequence and latent state given all the observations. This is achieved in this paper with an efficient backward sweep through the observations, in a new version of the method for standard HMMs described in \cite{Godsill2004}. Two new schemes are introduced: one for conditionally linear-Gaussian models which exploits the method of Rao-Blackwellisation; and a second for use with conditionally-deterministic models which uses an augmented target distribution in the style of an SMC sampler \cite{DelMoral2006}.

@@

We consider a general model from time $0$ to $T$, between which observations, $\{y_1 \dots y_N\}$, are made at times $\{t_1 \dots t_N = T\}$. During this period, an unknown number of changepoints, $K$, occur at times $\{\tau_0 = 0, \tau_1 \dots \tau_K \}$, each with associated changepoint parameters, $\{ u_0, u_1 \dots u_K \}$. The pairs $\{\tau_k, u_k\}$ are the elements of a marked point process (MPP). The latent state is a continuous-time process denoted $x(t)$. Discrete sets containing multiple values over time will be written as, e.g. $y_{1:n} = \{y_1 \dots y_n\}$.

@@

The changepoint sequence is assumed to be a Markov process,
%
\begin{IEEEeqnarray}{rCl}
 \{\tau_k, u_k\} & \sim & p(\tau_k, u_k|\tau_{k-1}, u_{k-1}) \label{eq:cp_model}     .
\end{IEEEeqnarray}

@@

This comprises a density term for each changepoint in the sequence and a survivor function term accounting for the probability that no other changepoints occur within the time interval.

@@

We first consider the class of variable rate models whose state dynamics are linear-Gaussian conditional on the changepoint sequence. Such a model may be discretised onto the set of observation times in exactly the same manner as a standard the standard HMM,
%
\begin{IEEEeqnarray}{rCl}
 x_n & = & A_n(\theta_{n})x_{n-1} + w_n \\
 y_n & = & C_n(\theta_{n})x_n + v_n       ,
\end{IEEEeqnarray}

where $w_n$ and $v_n$ are independent Gaussian variables (although the independence condition may be relaxed \cite{Grewal2002}),
%
\begin{IEEEeqnarray}{rCl}
 w_n & \sim & \mathcal{N}(w_n|0, Q_n(\theta_{n})) \\
 v_n & \sim & \mathcal{N}(v_n|0, R_n(\theta_{n}))       .
\end{IEEEeqnarray}

In addition, the prior state distribution should be Gaussian, with known mean and covariance,
%
\begin{IEEEeqnarray}{rCl}
 x_0 & \sim & \mathcal{N}(x_0|m_0, P_0)       .
\end{IEEEeqnarray}

If the changepoint sequence is known, or has been estimated, then the state values, $x_n$ may be inferred conditionally using optimal Kalman filtering and smoothing recursions. As well as the basic Kalman filter \cite{Kalman1960}, the Rauch-Tung-Striebel (RTS) smoother \cite{Rauch1965} and two-filter smoother \cite{Fraser1969} will prove useful for this step later in the paper.

Conditionally linear-Gaussian models for use with variable rate particle filters were introduced in \cite{Godsill2007a,Christensen2012} for financial and tracking problems. For example, changepoints might correspond to jumps in the value or trend of a security, at which points the process covariance, $Q_n(\theta_n)$, is inflated in an optimal fashion.

@@

In general, the state dynamics will be governed by a differential equation which may depend on the entire changepoint sequence. Here we assume that only the most recent changepoint determines the future state trajectory,
%
\begin{IEEEeqnarray}{rCl}
 dx(t) & = & \mathrm{f}(x(t), \tau_{K(t)}, u_{K(t)})     .
\end{IEEEeqnarray}

By introducing a new sequence, $\{ x_0, x_1 \dots x_K \}$, which denotes the value of the state at each changepoint (i.e. $x_k = x(\tau_k)$), and assuming that an analytic solution exists, a piecewise function describing the path may be found,
%
\begin{IEEEeqnarray}{rCll}
 x(t) & = & f(x_{K_n}, u_{K_n}, \tau_{K_n}, t) &, \tau_{K_n} < t \leq \tau_{K_{n}+1}    \label{eq:disc_time_state_diff_eq}     .
\end{IEEEeqnarray}

By choosing $t = \tau_{K_{n}+1}$, this equation specifies the state at the next changepoint time. Similarly, by choosing $t=t_n$, the state at the observation times may be calculated --- these latter points will be denoted $\hat{x}_n$. To complete the model, a probabilistic measurement model must be devised for the observation process, $p(y_n|\hat{x}_n)$.

@@

The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1})$, has a similar form to the changepoint prior of (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}, but an additional survivor function is included to account for the condition that changepoints cannot occur before $t_{n-1}$,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1}) = S(\tau_{K_n}, t_n) / S(\tau_{K_{n-1}+1}, t_{n-1}) } \nonumber \\
 \qquad  \qquad & & \times \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1})  \label{eq:cp_sequence_trandens}     .
\end{IEEEeqnarray}
%\begin{IEEEeqnarray}{rCl}
%\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1})} \nonumber \\
%    & = & p(\tau_{K_{n-1}+1:K_n}, u_{K_{n-1}+1:K_n}, \tau_{K_n+1}>t_n|\tau_{K_{n-1}+1}>t_{n-1}, \tau_{K_{n-1}}, u_{K_{n-1}}) \nonumber \\
%    & = & \begin{cases} S(\tau_{K_n}, t_n) \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1}, \tau_j>t_{n-1}) & K_n > K_{n-1} \\
%                        S(\tau_{K_n}, t_n) / S(\tau_{K_n}, t_{n-1}) & K_n = K_{n-1} \end{cases} \IEEEeqnarraynumspace \label{eq:cp_sequence_trandens}
%\end{IEEEeqnarray}

@@

Practically, because changepoints will be relatively rare events in most models, it is not likely that more than one new changepoint will occur between $t_{n-1}$ and $t_n$.

The target distribution of (\ref{eq:vrpf_target}) cannot be calculated analytically, but may be approximated numerically, for example, by using a particle filter.

A particle filter is an algorithm for approximating a probability distribution using a set of weighted samples (or ``particles'') drawn from that distribution. In this case, each particle will be a set of changepoint times and parameters.
%
\begin{equation}
 \hat{p}(\theta_{n}|y_{1:n}) = \sum_j w_n^{(j)} \delta_{\theta_{n}^{(j)}}(\theta_{n}) \label{eq:vrpf}
\end{equation}

@@

The particle filter works recursively. At the $n$th step, a particle, $\theta_{n-1}^{(i)}$, is first resampled from those approximating the filtering distribution at the $(n-1)$th step, using an appropriately chosen set of proposal weights, $\{v_{n-1}^{(j)}\}$ (where $\sum_j v_{n-1}^{(j)} = 1$),
%
\begin{equation}
 q(\theta_{n-1}) = \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})     .
\end{equation}

@@

Next, an extension to the changepoint sequence, $\theta_{n \setminus n-1}^{(i)}$, is sampled from an importance distribution, $q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n)$, and concatenated with $\theta_{n-1}$ to create a proposal for $\theta_n$. Finally, the particle is weighted according to the ratio of the target and proposal densities,
%
\begin{IEEEeqnarray}{rCl}
w_n^{(i)} & = & \frac{ p(\theta_{n}^{(i)}|y_{1:n}) }{ q(\theta_{n}) } \nonumber \\
    & \propto & \frac{ p(y_n|\theta_n, y_{1:n-1}) p(\theta_{n \setminus n-1}|\theta_{n-1}) p(\theta_{n-1}|y_{1:n-1}) }{ q(\theta_{n-1}) q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n) } \nonumber \\
    & =       & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times \frac{ p(y_n|\theta_n, y_{1:n-1}) p(\theta_{n \setminus n-1}|\theta_{n-1}) }{ q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n) } \label{eq:vrpf_weights}     .
\end{IEEEeqnarray}

@@

For the most basic ``bootstrap'' \cite{Gordon1993} form of the VRPF, $\theta_{n \setminus n-1}$ may be proposed from the prior transition density (\ref{eq:cp_sequence_trandens}). This can be achieved by sampling new changepoints sequentially from the transition model (\ref{eq:cp_model}) until one falls after the current time, $t_n$. This final future changepoint is discarded. (This process can be thought of as sampling the entire future changepoint sequence from $t_{n-1}$ onwards, and then marginalising those which fall after $t_n$.) The bootstrap proposal leads to the usual simplification of the weight formula, % (apart from the first which is sampled from (\ref{eq:cp_cond_model}))
%
\begin{IEEEeqnarray}{rCl}
w_n^{(i)} & = & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times p(y_n|\hat{x}_n) \label{eq:bootstrap_vrpf_weights}     .
\end{IEEEeqnarray}

@@

The bootstrap versions of the VRPF may perform poorly if changepoints are not obvious until significantly after they occur. For example, in a tracking example, if a jump occurs in the acceleration, but only the position is observed, then this change may not be obvious until a number of additional observations have arrived. In this case, the estimation may be improved by the introduction of resample-move (RM) steps \cite{Gilks2001}. In an RM scheme, optional Metropolis-Hastings (MH) moves are conducted to alter the particle states after the importance sampling has taken place. For variable rate models, any one of the previous changepoints, $\tau_k$, or associated parameters, $u_k$, could be adjusted. Because more observations are available than when the changepoint was first proposed, it may be possible to construct more informed proposals and so move the changepoints towards regions with higher posterior probability. It is even possible to retrospectively add or remove changepoints, using reversible jump MH moves \cite{Green1995}. Variable rate particle filters using RM with piecewise deterministic models are described in \cite{Whiteley2011,Gilholm2008}.

@@

A filter conducts inference sequentially as new observations are introduced. The purpose of a smoother is to produce a revised estimate once all the observations have been made, using future values to improve upon the filter performance. Estimating changepoints online is a challenging task because the presence of a change may not be obvious until after it has happened, as discussed above. Thus, a smoothing algorithm is expected to achieve improved performance in changepoint and state estimation.

The target distribution for a variable rate smoothing algorithm is the posterior distribution over the entire changepoint sequence, $p(\theta|y_{1:N})$. This distribution is approximated by the final step of the VRPF. However, in the same manner as the fixed rate filter-smoother of \cite{Kitagawa1996}, this approximation is likely to lack path-space diversity --- because of the necessary resampling step in the filtering algorithm, the particles all share the same set of changepoints before a particular time, with variation only appearing for changepoints closer to $T$. For a better characterisation of the smoothing distribution, it is necessary to rejuvenate the set of particles. This is achieved here with a backward pass through the observations in an extension of the forward-backward method described in \cite{Godsill2004}.

@@

It remains to devise a scheme for sampling the conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$. The simplifications used by the fixed rate particle smoother which exploit the Markovian nature of the state sequence \cite{Godsill2004} are of no help, because past changepoints are not independent of future observations in variable rate models (i.e. $p(\theta_{n}|\theta_{n}^{+}, y_{1:N}) \ne p(\theta_{n}|\theta_{n}^{+}, y_{1:n}))$. However, when the state dynamics are linear-Gaussian conditional on the changepoint sequence, it is possible split the dependence on the observations into past and future by introducing the current state as an additional variable. This device was used in \cite{Sarkka2012} in the derivation of the Rao-Blackwellised particle filter for fixed rate conditionally linear-Gaussian models,
%
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{p(\theta_{n}|\theta_{n}^+, y_{1:N})} \nonumber \\
\qquad & \propto & p(\theta_{n}, \theta_{n}^+| y_{1:N}) \nonumber  \\
       & =       & \int p(x_n, \theta_{n}, \theta_{n}^+| y_{1:N}) dx_n \nonumber  \\
       & \propto & \int p(y_{n+1:N}|x_n, \theta_{n}, \theta_{n}^+, y_{1:n}) p(x_n, \theta_{n}, \theta_{n}^+| y_{1:n}) dx_n \nonumber \\
       & = & \int p(y_{n+1:N}|x_n, \theta_{n}^+) p(x_n|\theta_{n}, y_{1:n}) dx_n \nonumber \\
       &   & \times p(\theta_{n}^+|\theta_{n}) p(\theta_{n}|y_{1:n})     .
\end{IEEEeqnarray}

Finally, the particle approximation of (\ref{eq:vrpf}) is substituted for the filtering distribution, $p(\theta_{n}|y_{1:n})$,
%
\begin{IEEEeqnarray}{rCl}
\hat{p}(\theta_{n}|\theta_{n}^+, y_{1:N}) & = & \sum_i \tilde{w}_{n}^{(i)} \delta_{\theta_{n}^{(i)}}(\theta_{n}) \label{eq:rb-vrps_back_cond}     ,
\end{IEEEeqnarray}

where the backwards conditional weights are given by,
%
\begin{IEEEeqnarray}{rCl}
 \tilde{w}_n & \propto & w_n \int p(y_{n+1:N}|x_n, \theta_{n}^+) \nonumber \\
             &         & \times p(x_n|\theta_{n}^{(i)}, y_{1:n}) dx_n p(\theta_{n}^+|\theta_{n}^{(i)}) \label{eq:rb-vrps_back_cond_weight}     .
\end{IEEEeqnarray}

@@

The integral in~(\ref{eq:rb-vrps_back_cond_weight}) contains two terms involving the current state, $x_n$. The first is the familiar state posterior generated by the Kalman filter using (\ref{eq:kf_predict_start})--(\ref{eq:kf_update_stop}),
%
\begin{IEEEeqnarray}{rCl}
p(x_n|\theta_{n}^{(i)}, y_{1:n}) & = & \mathcal{N}(x_n|m_n^{(i)}, P_n^{(i)})     .
\end{IEEEeqnarray}

@@

The second state-dependent term in (\ref{eq:rb-vrps_back_cond_weight}) is the likelihood, $p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)$. This may be considered to be an improper density over $x_n$, and may be calculated analytically using a backwards Kalman filter, in a similar manner to that used in the two-filter smoother \cite{Fraser1969,Kitagawa1994,Sarkka2012,Gelb1974}. Such a backwards Kalman filter uses the following recursions, which are derived in appendix~\ref{app:backwards_KF},
%
\begin{IEEEeqnarray}{rCl}
 p(y_{n+1:N}|x_n, \theta_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \\
 p(y_{n:N}|x_n, \theta_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n)
\end{IEEEeqnarray}
%
\begin{IEEEeqnarray}{rCl}
 \tilde{m}_n^- & = & A_{n+1}^{-1} \tilde{m}_{n+1} \label{eq:backward_kf_predict_start} \\
 \tilde{P}_n^- & = & A_{n+1}^{-1} (\tilde{P}_{n+1} + Q_{n+1}) A_{n+1}^{-T} \label{eq:backward_kf_predict_stop} \\
 \tilde{\mu}_n & = & C_n \tilde{m}_n^- \label{eq:backward_kf_update_start} \\
 \tilde{S}_n   & = & C_n \tilde{P}_n^- C_n^T + R_n \\
 \tilde{K}_n   & = & \tilde{P}_n^- C_n^T \tilde{S}_n^{-1} \\
 \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{K}_n (y_n - \tilde{\mu}_n) \\
 \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{K}_n \tilde{S}_n \tilde{K}_n^T \label{eq:backward_kf_update_stop}     .
\end{IEEEeqnarray}

@@

Substituting the Gaussian densities into (\ref{eq:rb-vrps_back_cond_weight}) and performing the integral yields the following backwards conditional weights,
%
\begin{equation}
 \tilde{w}_n \propto w_n p(\theta_{n}^+|\theta_{n}^{(i)}) \mathcal{N}(\tilde{m}_n^-|m_n, \tilde{P}_n^- + P_n)     .
\label{eq:rb-vrps_back_cond_weight2}
\end{equation}

Using these weights, a sample of $\theta_{n}$ may be drawn from the particle distribution of (\ref{eq:rb-vrps_back_cond}), completing one step of the smoothing algorithm. Once sampling has progressed backwards from $n=N \dots 1$, a particle from the smoothing distribution will have been generated. This procedure may then be repeated until sufficient particles have been obtained. The algorithm is summarised in Fig.~\ref{alg:RBVRPS}.

@@

In a conditionally deterministic system, the entire continuous-time state trajectory, $x(t)$, is a function of the changepoint sequence $\theta$ and the initial state, $x_0$. When the changepoint sequence is split into past, $\theta_n$, and future $\theta_n^+$, and the past is altered, this will alter the state trajectory in the future. This is undesirable in practice, as each sampling operation will require the recalculation of state values and observations likelihoods for all future observation times.
